Refinement for open automata

### 2021-05-15 2021-Refinement/report.tex
Checked up to section 5

\usepackage{xunicode}
\usepackage{fontspec}
\usepackage[hmargin=0.5in,marginparwidth=1.5in,includemp]{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{unicode-math}
\usepackage{tikz}
\usetikzlibrary{external}
% \usepackage{hyperref}
% \usepackage{xcolor}
\AtBeginDocument{\renewcommand\setminus{\smallsetminus}}
\usepackage{polyglossia}
\setmainlanguage{english}

@@

% possible examples for refinement: + <= |, ; <= |
% transitions between sections

@@

\section{Introduction}
The model of open automata is used to give the semantic of open pNets \cite{2007.10770}.
Open pNets are not petri nets but an intermediate representation used in the VerCors project to perform verification and model checking \cite{henrio:01252323}.
Open pNets with the open automaton semantic and equivalence relation introduced in previous work \cite{2007.10770} were sucessfully used to verify BIP and GCM \cite{qin:01823507, ameurboulifa:01526055}.
However equivalences relations are not sufficient to perform model checking, refinement relations are also used.
For instance an example of open pNet where equivalence is not sufficient was encountered in ...\marginpar{Paper where the motivating example came up}.

In this article we introduce a refinement relation for open automata\marginpar{TODO: expand}.

We begin in section \ref{sec:notations} by giving some notations used throughout this paper.
Then there are several sections dedicated to define all the interesting objects and properties:
In section \ref{sec:def} we give a clean and up to date definition of open automata.
In section \ref{sec:semcompo} we recall the semantic for open automata and their composition without requiring the use of open pNets.
In section \ref{sec:proofelts} we adapt the standard properties related to refinement relation from LTS to open automata.
After that the section \ref{sec:prelref} is dedicated to progressively build a refinement relation.
Finally the section \ref{sec:refinement} introduces and analyses the refinement relation on open automata.

@@

Throughout this paper, tuples might be noted differently depending on what they represent.

Family of values, or equivalently maps will be noted \(\mset{i \mapsto x_i}{i \in I}\), \(\mset{i \gets x_i}{i \in I}\) or \(x_i^{i \in I}\).
The latter will only be used when there is a generating expression; for instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\).
However \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) has no generating expression and is represented with the first notation in a finite context.
The disjoint union of two maps \(\varphi: I \to X\) and \(\psi: J \to Y\) with \(I \cap J = \emptyset\) is \(\varphi \uplus \psi: I \uplus J \to X \cup Y\).

@@

\section{Open Automaton}\label{sec:def}
To define an open automaton we need some preliminary definitions.
\begin{defi}[Expression algebra, Action algebra, Formulae, Terms]
An expression algebra \(E\) is a disjoint union \(E \defobject \terms \uplus \actions \uplus \formulae\) of the terms, the actions and the formulae.

The terms \(\terms\) is a term algebra.
The action algebra \(\actions\) is the set of pairs \(\labels \times \terms\) where \(\labels\) are the action labels.
The formulae \(\formulae\) are at least the first order formulae over \(\terms\) and \(\labels\).
\end{defi}
The term algebra and action labels are arbitrary.
An element \(\mpar{l, t} \in \actions\) is noted \(l\mpar{t}\).
The formulae contain at least first order logic terms with an equality relation.
This equality relation is not necessarily a syntactic equality (\(2 + 2 \neq 4\) with a syntactic equality) on \(\terms\).

An example of term algebra can be Peano integers (zero, variable and successor function), the formulae associated can use equality relation, the sum relation \(\mathit{sum}\mpar{a, b, c} \defobject a = b + c\) and the product relation \(\mathit{prod}\mpar{a, b, c} \defobject a = b \times c\).

\begin{defi}[Free variables, Closed terms, Closed formulae]
\defitem \(\fvars{e}\) is the set of unbound variables in \(e \in E\).
\defitem The expressions restricted to variables in \(V\) are \(E_V \defnotation \mset{e \in E}{\fvars{e} \subseteq V}\); \(E_V \subset E\).
\defitem The closed expressions are expressions restricted to variables in \(\emptyset\), \(E_\emptyset\).
\end{defi}
Terms, actions and formulae restricted to variables and their closed versions are also defined by restriction.
A closed expression can contain variables, only unbound variable are restricted.

We can use the previous example of first order formula on Peano integers to illustrate these definitions.
\(x\), \(S\mpar{y}\) are valid terms, \(0\) is a closed term, \(\fvars{x} = \mbrc{x}\), \(\fvars{S\mpar{y}} = \mbrc{y}\), \(\fvars{0} = \emptyset\).
\(x = 0\), \(\forall y, \exists x, \mathit{sum}\mpar{y, z, x}\) are valid formulae, \(\forall x, \neg S\mpar{x} = 0\) is a valid closed formula, \(\fvars{x = 0} = \mbrc{x}\), \(\fvars{\forall y, \exists x, \mathit{sum}\mpar{y, z, x}} = \mbrc{z}\), \(\fvars{\forall x, \neg S\mpar{x} = 0} = \emptyset\).

\begin{defi}[Interpretation, Values, Satisfiability, (Parallel) substitution]
We assume that the following are given:
\defitem The values \(\values\), which are interpretations of closed terms.
\defitem The satisfiability relation on closed formulae, \({\vdash} \subseteq \rformulae\).
\defitem The substitution in \(e \in E\) of \(x \in \fvars{e}\) by \(t \in \terms\), \(e\subst{t}{x}\).
\defitem The parallel substitution in \(e \in E\) of variables in \(V\) by \(\psi: V \to \terms\), \(e\psubst{\psi}\).
\end{defi}
For the parallel substitution, the set \(V\) is not required to be a subset of \(\fvars{e}\).
In the case where it is not, the variables in \(V \setminus \fvars{e}\) are not substituted.
The substitutions might give a ill-formed expression; for instance let the terms be integers and pairs with (pointwise) addition, \(\mpar{a + b}\psubst{a \mapsto 7, b \mapsto \mpar{4, 5}}\) is a ill-formed term.
This can be guarded with the check \(e\subst{t}{x} \in E\) and \(e\psubst{\psi} \in E\) and it will implicitly be the case, for instance when there is quantification on \(t\) and \(\psi\), to simplify notations.

The interpretation of terms is supposed to be decidable.
The satisfiability of formulae might not be decidable nor complete nor consistent, however we will pretend like they are because these are really hard problems for logicians that we don't want to deal with.
For instance a formula with quantifiers on variables might not be provable even if it is true for all values of these variables.
In practise the formulae will be given to a SMT solver and we make sure they have all the previous properties.
Actually \(\vdash\) is used to separate the external logic and the logic on \(\formulae\).

Values will be used for keeping a variable state, and then injected in terms for substitution.
This is correct when \(\values \subseteq \rterms\) and we suppose it is the case.
Otherwise it doesn't invalidates theorem because it can stand as a shorthand for substitution with any term which is interpreted as \(v\).
We suppose that the interpretation of terms is compatible w.r.t.\@ substitution, that is if two terms \(t\), \(t'\) are interpreted with the same value, then replacing \(t\) by \(t'\) in a well-formed expression makes an equivalent well-formed expression.
\begin{noti}[Notations for separating external logic and logic on \(\formulae\)]
\defitem The satisfiability relation will be noted using the classical relational notation \(\vdash f\).
\defitem The satisfiability of a formula \(f \in \formulae\) under some valuation \(\sigma: V \to \values\) is noted:
\[ \sigma \vdash f \defnotation \vdash \exists \fvars{f\psubst{\sigma}}, f\psubst{\sigma} \]
\defitem The satisfiability of a formula \(f \in \formulae\) with some variable set \(V\) as context is noted:
\[ V \vdash f \defnotation \vdash \forall V, \exists\mpar{\fvars{f} \setminus V}, f \]
\defitem The precedence of \(\vdash\) is the lowest on the right side and higher than \(\uplus\) on the left side:
\[ \forall a \, b, a \uplus b \vdash x \wedge y \implies \exists z, P\mpar{x,z} \text{ is the same as } \forall a \, b, \mpar{\mpar{a \uplus b} \vdash \mpar{\mpar{x \wedge y} \implies \exists z, P\mpar{x,z}}} \] % Mettre des niveaux de parenthèses
\end{noti}
With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
A open automaton is a tuple \(\OA{S}{s_0}{V}{\sigma_0}{J}{T}\) with \(S\) the set of states, \(s_0 \in S\) the initial state, \(V\) the set of variable names unique to this automaton, \(\sigma_0: V' \to \values\) the initial valuation of variables where \(V' \subseteq V\), \(J\) the set of hole names and \(T\) the set of open transitions.

\(S, V, J\) are arbitrary finite sets.
\end{defi}
The variable names may clash when considering two automata, in that case we suppose that we can still distinguish the variables in the formulas.
% The open transitions will be considered to be a finite set of representative open transitions. Removing this restriction may break some theorems.
The initial valuation may be a partial valuation of the variables, if the undefined variables are set before being read this behave like a normal variable.
If the undefined variable is not set before being read then its value may be any fixed value, the potential use of this construct will be explain in more details in Section \ref{sec:prelref}.
\begin{defi}[Open transition]
An open transition is a tuple \nmm{\OTg} with \(s, s' \in S\) the source and target states, \(\alpha \in \actions\) the produced action, \(J' \subseteq J\) the holes involved, \(\beta_j \in \actions\) the actions of the holes, \(g \in \rformulae\) the guard and \(\psi: V \to \terms\) the new value of the variables.
\end{defi}
An open transition can have many unbound variable, actually an effective transition of the automaton is any well-formed substitution of the unbound variables of the transition minus the automaton variables.
% This is used to constrain the infinite branching to be representable, making it possible to reason about.

The intuition of a semantic open automaton is a partially defined LTS with variables, guards on transitions and parametrised actions.
Initially the automaton is in the initial state with the initial valuation.
In any state with any valuation, it can perform transitions which source state is the current state and which guard is valid in the current valuation if the holes emit the indicated action.
By performing the transition the automaton emits the indicated action and updates its variables and state.

\begin{noti}[Bisimilarity, FH-bisimulation]
The bisimilarity relation on open automata is noted \(\cong\) and the FH-bisimulation \cite{henrio:01055091} is noted \(\overset{FH}=\).
\end{noti}
Note that FH-bisimilarity implies bisimilarity.

To illustrate these definitions we consider  two implementations (figure \ref{fig:enable}) of the LOTOS \cite{ISOLOTOS} operator enable in the open automata model.
The enable operator runs  the left hand side agent until it chooses to finish, at which point it produces an action \(\delta\) that is synchronised with the first action of the right hand side agent which must be \(\act{acc}\), then only the latter agent runs.
During the synchronised action \Quentin{data can be exchanged}{I need to read a little bit more on that; is that a return code?}.

\begin{exi}[Enable, state-oriented]
\begin{figure}
\centering
\input{enable_state.tex}
\vrule
\input{enable_var.tex}
\caption{Enable operator implementation with open automata, on the left state oriented, on the right variable oriented}
\label{fig:enable}
\end{figure}

Graphical notations for showing automata are as follows.
Automata are shown in the figure, the usual circle for states and simple arrows for transitions convention is used.
The initial state is indicated by a double circle.
States names are indicated inside the circles and transitions labels are drawn near their corresponding arrow.
The open transitions do not indicate the source and target states since that is the role of the transitions arrows; only the action is on the bottom side of the open transition.
Also in the example, transitions are specified using variables in place of actions.
This is used to represent all the transitions created by a correct substitution of these variables, which cannot be drawn as there are often infinitely many.

We can proceed to explain the left enable operator example.
The automata on the left side is \(\OA{\mbrc{L, R}}{L}{\emptyset}{\mbrc{}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{L}{L}{x}{\mbrc{l \mapsto x}}{x \neq \delta}{\mbrc{}} &&
	\OT{L}{R}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{x = y}{\mbrc{}} &&
	\OT{R}{R}{x}{\mbrc{r \mapsto x}}{\top}{\mbrc{}}
\end{align*}
Note that we could have expressed the transition in the middle with \(\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{x}}\) which would have avoided many transitions with trivially false guards like the one where \(x \mapsto 1, y \mapsto 2\) which has the guard \(1 = 2\).

This automaton is an implementation of the enable operator because it begins in the state \(L\) where it allows any transition from its hole \(l\) which is not \(\delta\) then the automaton synchronises its holes on the same data, effectively allowing a data exchange when it goes into state \(R\), and finally allows any transition from its hole \(r\).
We use the standard convention that uses \(\tau\) as a non-observable transition, never synchronised with other actions and passed unmodified.
This allows the synchronisation of the two holes to be hidden to the exterior by sending a \(\tau\).
However here \(\tau\) is not always allowed from the holes, for instance in the state \(L\) the hole \(r\) cannot emit it.
These transition have been omitted for the sake of simplifying the first example of open automaton.
\end{exi}

Finally we can define some utilitary functions:
\begin{defi}[Guard, Out-transition, Transition variables]
Let \(V\) be the variable names of the considered automaton, \(T\) its transitions and \(s\) be one of its states.
\(\fOT{s}\) are called the out-transitions of \(s\).
\(\fIT{s'}\) are called the in-transitions of \(s'\).
\begin{align*}
	\fOT{t} & \defnotation \mset{\OTg \in T}{s = t} &
	\fIT{t} & \defnotation \mset{\OTg \in T}{s' = t} \\
\end{align*}
\vspace{-1cm}
\begin{gather*}
	\fguard{\OTg} \defnotation g \\
	\fvars{\OTg} \defnotation \mpar{\fvars{\alpha} \cup \fvars{g} \cup \bigcup_{j \in J'} \fvars{\beta_j} \cup \bigcup_{v \in V} \fvars{\psi\mpar{v}}} \setminus V
\end{gather*}
\end{defi}

\begin{exi}[Enable, variable-oriented]
The automata drawn on the right side of the figure is an alternative implementation of the enable operator which is FH-bisimilar to the previous one.
The drawing introduces the double arrows which is used to denote the initial valuation.
The automata drawn is \(\OA{\mbrc{\_}}{\_}{\mbrc{v}}{l^\mbrc{v}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{\_}{\_}{x}{\mbrc{l \mapsto x}}{v = l \wedge x \neq \delta}{\mbrc{}} &&
	\OT{\_}{\_}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{v = l \wedge x = y}{\mbrc{v \gets r}} &&
	\OT{\_}{\_}{x}{\mbrc{r \mapsto x}}{v = r}{\mbrc{}}
\end{align*}
Note that in this example \(r\) and \(l\) are a hole names and also closed terms: this is a naming conflict, we do not have \(J \subseteq \rterms\).

The variables of a transition are all the free variables in the expressions minus the automaton variables.
For instance \(v\) is a variable name for the automaton which is used in the guard.
It is not a transituon variable.
When generating transition without free variables its value is not substituted with a closed term.
For instance \nmm{\OT{\_}{\_}{\tau}{\mbrc{l \mapsto \tau}}{v = l \wedge \tau \neq \delta}{\mbrc{}}} is a transition generated by the second transition and is in \(T\) but not \nmm{\OT{\_}{\_}{\tau}{\mbrc{l \mapsto \tau}}{l = l \wedge \tau \neq \delta}{\mbrc{}}} (result of the forbidden substitution \(v \mapsto l\)).

An imaginary run of this open automaton can be the automaton in the hole \(l\) emits many actions, this synchronises with the only transition possible at that moment because \(v = l\).
Then the automaton in hole \(l\) emits a \(\delta\mpar{t}\), this is synchonised with the action that sets \(v \gets r\) and the automata in hole \(r\) must emit a \(\act{acc}\mpar{t}\).
Finally the automaton in the hole \(r\) emits any sequence of action.

Another imaginary run can be the automaton in the hole \(l\) emits a sequence of action and the automaton in the hole \(r\) cannot emit a \(\act{acc}\mpar{t}\) so it never runs.
\end{exi}

From this point open automata and open transitions will be called automata and transitions for simplicity.

>>>>

\usepackage{xunicode}
\usepackage{fontspec}
\usepackage[hmargin=0.5in,marginparwidth=1.5in,includemp]{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{unicode-math}
\usepackage{tikz}
\usetikzlibrary{external}
\usepackage[colorlinks]{hyperref}
% \usepackage{xcolor}
\AtBeginDocument{\renewcommand\setminus{\smallsetminus}}
\usepackage{polyglossia}
\setmainlanguage{english}

@@

% possible examples for refinement: + <= |, ; <= |
% transitions between sections
% TODO: CITER LES CONFERENCES

@@

\section{Introduction}
The open automata model is used to give the semantic of open pNets \cite{2007.10770}.
Open pNets are not petri nets but an intermediate representation used in the VerCors project to perform verification \cite{henrio:01252323}.
Verification is the process of checking that an implementation satisfies a specification.
Open pNets with the open automaton semantics and equivalence relation introduced in previous work \cite{2007.10770} were sucessfully used to model and verify BIP and GCM \cite{qin:01823507, ameurboulifa:01526055}.
However equivalences relations are not always sufficient to perform verification, refinement relations are also used.
For instance an example of open pNet where equivalence is not sufficient was encountered in ...\marginpar{Paper where the motivating example came up}.

The contributions of this article are the definition of composition for open automata without having to define open pNets, and the introduciton of several refinement relations for open automata\marginpar{TODO: expand}.

We begin with Section \ref{sec:notations} by giving some notations used throughout this paper.
Then there are several sections dedicated to define all the interesting objects and properties:
In Section \ref{sec:def} we give a clean and up to date definition of open automata.
In Section \ref{sec:comp} we define their composition without requiring the use of open pNets.
In Section \ref{sec:proofelts} we adapt the standard properties related to refinement relation from LTS to open automata.
After that Section \ref{sec:prelref} is dedicated to progressively build a refinement relation.
Section \ref{sec:refinement} introduces and analyses the refinement relation on open automata.
% flat model
% future work
% conclusion

@@

Throughout this paper, tuples will be noted differently depending on what they represent.
This helps distinguishing the manipulated objects.
Every such notation will be introduced in the definition of the object.

Families of values, or equivalently maps will be noted \(\mset{i \mapsto x_i}{i \in I}\), \(\mset{i \gets x_i}{i \in I}\) or \(x_i^{i \in I}\).
The latter can only be used if there is a generating expression; for instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\).
However \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) has no generating expression and is represented here with the finite version of first notation.
The disjoint union of two maps \(\varphi: I \to X\) and \(\psi: J \to Y\) with \(I \cap J = \emptyset\) is \(\varphi \uplus \psi: I \uplus J \to X \cup Y\).

@@

\section{Open Automata}\label{sec:def}
To define the open automata we need some preliminary definitions.
\begin{defi}[Expression algebra, Action algebra, Formulas, Terms]
An expression algebra \(E\) is a disjoint union \(E \defobject \terms \uplus \actions \uplus \formulas\) of the terms, the actions and the formulas.

The terms \(\terms\) is a term algebra\marginpar{reference to come}.
As any term algebra it has constant symbols with arity, variables, and a typing mechanism to distinguish well-formed and ill-formed terms.
The action algebra \(\actions\) is another term algebra.
It can be a subset of the terms.
The formulas \(\formulas\) are at least the first order formulas over \(\terms\) and \(\actions\).
\end{defi}
The term algebras are arbitrary.
The formulas contain at least first order logic terms with an equality relation.
This equality relation is not necessarily a syntactic equality (\(2 + 2 \neq 4\) with a syntactic equality).

An example of term algebra can be Peano integers (constant zero with arity 0, constant successor function with arity 1 and the variables), the formulas associated can use syntactic equality relation, the sum relation \(\mathit{sum}\mpar{a, b, c}: \text{``}a = b + c\text{"}\) and the product relation \(\mathit{prod}\mpar{a, b, c}: \text{``}a = b \times c\text{"}\).

\begin{defi}[Unbound variables, Expressions restricted to variables, Closed expressions]
\defitem \(\fvars{e}\) is the set of variables in \(e \in E\) that are not bound by any binder.
	Binders can be for instance quantifiers in formulas, or let-binders in terms if they are part of the term algebra.
\defitem The expressions restricted to variables in \(V\) are \(E_V \defnotation \mset{e \in E}{\fvars{e} \subseteq V}\); \(E_V \subset E\).
\defitem The closed expressions are expressions restricted to variables in \(\emptyset\), \(E_\emptyset\).
\end{defi}
Terms, actions and formulas restricted to variables and their closed versions are also defined by restriction.
A closed expression can contain variables under a binder, only unbound variable are forbidden.

We can use the previous example of first order formula on Peano integers to illustrate these definitions.
\(x\), \(S\mpar{y}\) are well-formed terms, \(0\) is a closed term, \(\fvars{x} = \mbrc{x}\), \(\fvars{S\mpar{y}} = \mbrc{y}\), \(\fvars{0} = \emptyset\).
\(x = 0\), \(\forall y, \exists x, \mathit{sum}\mpar{y, z, x}\) are valid formulas, \(\forall x, \neg S\mpar{x} = 0\) is a valid closed formula, \(\fvars{x = 0} = \mbrc{x}\), \(\fvars{\forall y, \exists x, \mathit{sum}\mpar{y, z, x}} = \mbrc{z}\), \(\fvars{\forall x, \neg S\mpar{x} = 0} = \emptyset\).

\begin{defi}[Values, Satisfiability, (Parallel) substitution]
We assume that the following are given:
\defitem The values \(\values\), which are interpretations of closed terms.
\defitem The satisfiability relation on closed formulas, \({\vdash} f\) where \(f \in \rformulas\).
\defitem The substitution in \(e \in E\) of \(x \in \fvars{e}\) by \(t \in \terms\), \(e\subst{t}{x}\).
\defitem The parallel substitution in \(e \in E\) of variables in \(V\) by \(\psi: V \to \terms\), \(e\psubst{\psi}\).
\end{defi}
For the parallel substitution, the set \(V\) is not required to be a subset of \(\fvars{e}\).
In the case it isn't, the variables in \(V \setminus \fvars{e}\) are not substituted.
The substitutions might give a ill-formed expression; for instance let the terms be integers and pairs with (pointwise) addition, \(\mpar{a + b}\psubst{a \mapsto 7, b \mapsto \mpar{4, 5}}\) is a ill-formed term.
This can be guarded with the check \(e\subst{t}{x} \in E\) and \(e\psubst{\psi} \in E\) and it will implicitly be the case, for instance when there is quantification on \(t\) and \(\psi\), to simplify notations.

The interpretation of terms is supposed to be decidable.
The satisfiability of formulas might not be decidable nor complete nor consistent, however we will pretend like they are because these are really hard problems for logicians that we don't want to deal with.
For instance a formula with quantifiers on variables might not be provable even if it is true for all values of these variables.
In practise the formulas will be given to a SMT solver and we cannot always make sure they have all the previous properties.
\(\vdash\) can hence be interpreted as an indicator of what is given to the SMT; it separates the external logic and the logic on \(\formulas\).

Values will be used for keeping a variable state, and then injected in terms for substitution.
This is correct when \(\values \subseteq \rterms\) and we suppose it is the case.
Otherwise it doesn't invalidates theorems because it can stand as a shorthand for substitution with any term which is interpreted the kept value.
We suppose that the interpretation of terms is compatible w.r.t.\@ substitution, that is if two terms \(t\), \(t'\) are interpreted with the same value, then replacing \(t\) by \(t'\) in a well-formed expression makes an equivalent well-formed expression.
\begin{noti}[Notations for separating external logic and logic on \(\formulas\)]
\defitem The satisfiability of a formula \(f \in \formulas\) under some valuation \(\sigma: V \to \values\) is noted:
\[ \sigma \vdash f \defnotation \vdash \exists \fvars{f\psubst{\sigma}}, f\psubst{\sigma} \]
\defitem The satisfiability of a formula \(f \in \formulas\) with some variable set \(V\) as context is noted:
\[ V \vdash f \defnotation \vdash \forall V, \exists\mpar{\fvars{f} \setminus V}, f \]
\defitem The precedence of \(\vdash\) is the lowest on the right side and higher than \(\uplus\) on the left side:
\[ \forall a \, b, a \uplus b \vdash x \wedge y \implies \exists z, P\mpar{x,z} \text{ is the same as } \forall a \, b, \bigg((a \uplus b) \vdash \Big((x \wedge y) \implies \exists z, P\mpar{x,z}\Big)\bigg) \]
\end{noti}
With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
A open automaton is a tuple \(\OA{S}{s_0}{V}{\sigma_0}{J}{T}\) with \(S\) the set of states, \(s_0 \in S\) the initial state, \(V\) the set of variable names unique to this automaton, \(\sigma_0: V' \to \values\) the initial valuation of variables where \(V' \subseteq V\), \(J\) the set of hole names and \(T\) the set of open transitions.

\(S, V, J\) are arbitrary finite sets. % OR finite arbitrary sets?
\end{defi}
The variable names may clash when considering two automata, in this case we suppose that we can still distinguish the variables in the formulas.
In practise the open automata are used in a toolchain at a point where what came before guarantees there are no name clash, which is why we make this assumption.

The initial valuation may be a partial valuation of the variables.
If an undefined variable is set before being read, it behaves like an initially set variable.
If an undefined variable is not set before being read then its value may be any fixed value.
When refining the automaton it can be set by the implementation, or it can be left unspecified as a parameter of the automaton (for instance a \(n\) bits register has \(n\) as parameter).
\begin{defi}[Open transition]
An open transition is a tuple \nmm{\OTg} with \(s, s' \in S\) the source and target states, \(\alpha \in \actions\) the produced action, \(J' \subseteq J\) the holes involved in the transition, \(\beta_j \in \actions\) the actions of the holes, \(g \in \formulas\) the guard and \(\psi: V \to \terms\) the variables assignements.
\end{defi}
An open transition can have many unbound variables.
Actually an effective transition of the automaton is any well-formed substitution of the unbound variables of the transition minus the automaton variables.

The intuition of an open automaton is a partially defined LTS with variables, guards on transitions and parametrised actions.
Initially the automaton is in the initial state with the initial valuation.
In any state with any valuation, it can perform effective transitions which source state is the current state and which guard is satisfiable in the current valuation if the holes emit the indicated actions.
An effective transition is a transition where every variable that is not an automaton variable is instanciated with a value.
By performing the effective transition, the automaton emits the indicated action and updates its variables and state according to the target state and variables assignements.

\begin{noti}[FH-bisimulation]
The FH-bisimulation \cite{henrio:01055091} is noted \(\cong\).

The FH-bisimulation is currently the only equivalence relation on open automata.
\end{noti}
To illustrate these definitions we consider two implementations (figure \ref{fig:enable}) of the LOTOS \cite{ISOLOTOS} operator enable in the open automata model.
The enable operator runs  the left hand side agent until it chooses to finish, at which point it produces an action \(\delta\mpar{t}\) (with \(t\) some data) that is synchronised with the first action of the right hand side agent which must be \(\act{accept}\mpar{t}\) (here \(t\) is an input), then only the latter agent runs.
The \(\act{accept}\) action is shortened as \(\act{acc}\) in the examples.
During the synchronised action \Quentin{data can be exchanged}{I need to read a little bit more on that; is that a return code?}.
\begin{exi}[Enable, state-oriented]
Graphical convention for drawing automata are as follows.
As standard automata, circles represent states and simple arrows represent transitions.
The initial state is indicated by a double circle.
States names are indicated inside the circles and transitions labels are drawn near their corresponding arrow.
The open transitions do not indicate the source and target states since that is the role of the transitions arrows; only the emmited action is on the bottom side of open transitions.
Initial valuations are indicated near a double lined arrow pointing to the initial state.

\begin{figure}
\centering
\input{enable_state.tex}
\vrule
\input{enable_var.tex}
\caption{Enable operator implementation with open automata, on the left state oriented, on the right data oriented}
\label{fig:enable}
\end{figure}
The automaton on the left side of the figure is \(\OA{\mbrc{L, R}}{L}{\emptyset}{\mbrc{}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{L}{L}{x}{\mbrc{l \mapsto x}}{\forall y, x \neq \delta\mpar{y}}{\mbrc{}} &&
	\OT{L}{R}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{x = y}{\mbrc{}} &&
	\OT{R}{R}{x}{\mbrc{r \mapsto x}}{\top}{\mbrc{}}
\end{align*}
Note that the transition in the middle could have been expressed as \(\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{x}}\).
It would have avoided many effective transitions with trivially false guards like the one where \(x \mapsto 1, y \mapsto 2\), which has the guard \(1 = 2\).

This automaton is an implementation of the enable operator because it begins in the state \(L\), where it allows any non \(\delta\) transition from its hole \(l\), then the automaton synchronises its holes on the same data, effectively allowing a data exchange when it goes into state \(R\), and finally allows any transition from its hole \(r\).
The implementation of value passing in the open automata model doesn't distinguish input and output variables as in LOTOS.
Both holes must emit their actions with valid data to perform the transition in the enable operator.
If it is not possible the system is locked.

We use the standard convention that uses \(\tau\) as a non-observable transition, never synchronised with other actions and passed unmodified.
This allows the synchronisation of the two holes to be hidden to the exterior by sending a \(\tau\).
However here \(\tau\) is not always allowed from the holes, for instance in the state \(L\) the hole \(r\) cannot emit it.
These transitions have been omitted for the sake of simplifying the first example of open automata.
\end{exi}
Finally we can define some utilitary functions:
\begin{defi}[Guard, Out-transition, Transition variables]
Let \(V\) be the variable names of the considered automaton, \(T\) its transitions and \(r\) one of its states.
\(\fOT{r}\) are called the out-transitions of \(r\).
\(\fIT{r}\) are called the in-transitions of \(r\).
The local variables of a transition are all variables appearing in that transition except the global variables of the automaton.
\begin{align*}
	\fOT{r} & \defnotation \mset{\OTg \in T}{s = r} &
	\fIT{r} & \defnotation \mset{\OTg \in T}{s' = r} \\
\end{align*}
\vspace{-1cm}
\begin{gather*}
	\fguard{\OTg} \defnotation g \\
	\fvars{\OTg} \defnotation \mpar{\fvars{\alpha} \cup \fvars{g} \cup \bigcup_{j \in J'} \fvars{\beta_j} \cup \bigcup_{v \in V} \fvars{\psi\mpar{v}}} \setminus V
\end{gather*}
\end{defi}
The goal of the extractor \(\fvars{t}\) when \(t\) is a transition is to get variables that are unique to that transition.
Other variables like automaton variables are already known at this point so there is no use in getting them.
Also the use of this extractor benefits from this exclusion: otherwise there would always be \(\setminus V\) following it to prevent variable shadowing.

\begin{exi}[Enable, variable-oriented]
The automaton drawn on the right side of Figure \ref{fig:enable} is \(\OA{\mbrc{\_}}{\_}{\mbrc{v}}{l^\mbrc{v}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{\_}{\_}{x}{\mbrc{l \mapsto x}}{v = l \wedge x \neq \delta}{\mbrc{}} &&
	\OT{\_}{\_}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{v = l \wedge x = y}{\mbrc{v \gets r}} &&
	\OT{\_}{\_}{x}{\mbrc{r \mapsto x}}{v = r}{\mbrc{}}
\end{align*}
It is an alternative implementation of the enable operator which is FH-bisimilar to the previous one, as proved in \marginpar{probably forte, check that}.
Note that in this example \(r\) and \(l\) are a hole names and also closed terms: this is a naming conflict, though it is not ambiguous because \(J \cap \rterms = \emptyset\) always holds.
The variables local to a transition are all the unbound variables in the expressions minus the automaton variables.
For instance \(v\) is a variable of the automaton (used in the guard).
It is not local to the transition.
In a run, when taking effective transitions (without local variables), its value is not substituted with a closed term.
For instance \nmm{\OT{\_}{\_}{\tau}{\mbrc{l \mapsto \tau}}{v = l \wedge \tau \neq \delta}{\mbrc{}}} is an effective transition generated from the second transition but not \nmm{\OT{\_}{\_}{\tau}{\mbrc{l \mapsto \tau}}{l = l \wedge \tau \neq \delta}{\mbrc{}}} (result of the forbidden substitution \(v \mapsto l\)).

An imaginary run of this open automaton can be: The automaton in the hole \(l\) emits many actions, this synchronises with the only transition possible at that moment because \(v = l\).
Then the automaton in hole \(l\) emits a \(\delta\mpar{t}\), this is synchonised with the action that sets \(v \gets r\) and the automata in hole \(r\) must emit a \(\act{acc}\mpar{t}\).
Finally the automaton in the hole \(r\) emits any sequence of actions.

Another imaginary run can be: The automaton in the hole \(l\) emits a sequence of non \(\delta\mpar{t}\) actions so the automaton in the hole \(r\) cannot emit a \(\act{acc}\mpar{t}\) and never runs.
\end{exi}

From this point open automata and open transitions can be called automata and transitions for simplicity. % NEED NORMALISATION, LEFT IN CASE OF OMISSION
